---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

Load required packages
```{r}
library(knitr)
library(tidyverse)
library(lattice)
```


Set default to be echo = TRUE
```{r}
opts_chunk$set(echo = TRUE)
```

Set format of numbers
```{r}
options(digits = 2, scipen = 10)
```

## Loading and preprocessing the data

Unzip the folder containing the data
```{r}
unzip("activity.zip")
```

Read in the data
```{r}
activity_data <- read_csv("activity.csv")
```

View the first few rows of the data
```{r}
head(activity_data)
```

## What is mean total number of steps taken per day?

Group the data by day. Find the total steps taken on each day (excluding missing values)
```{r}
steps_per_day <- activity_data %>%
  na.omit() %>%
  group_by(date) %>%
  summarise(total_steps = sum(steps))
steps_per_day
```

Construct a histogram of the total number of steps taken each day
```{r histogram}
hist(steps_per_day$total_steps,
     col = "green",
     main = "Total Number of Steps Taken per Day",
     xlab = "Number of steps",
     ylab = "Frequency")
```

Calculate the mean and median number of steps taken in a day
```{r}
mean_steps <- mean(steps_per_day$total_steps)
median_steps <- median(steps_per_day$total_steps)
```

The mean number of steps taken per day is `r mean_steps` steps.  
The median number of steps taken per day is `r median_steps` steps.

## What is the average daily activity pattern?

Find the average number of steps taken at each interval (exluding missing values)
```{r}
intervals <- activity_data %>%
  na.omit() %>%
  group_by(interval) %>%
  summarise(avg_steps = mean(steps))
intervals
```

Construct a time series plot of the 5-minute intervals and average number of steps taken
```{r timeplot}
plot(x = intervals$interval, y = intervals$avg_steps,
     type = "l",
     col = "blue",
     main = "Average Number of Steps Taken during each Interval",
     xlab = "Interval",
     ylab = "Number of steps")
```

Find the interval which contains the maximum number of steps on average
```{r}
max_steps <- intervals %>% filter(avg_steps == max(avg_steps))
max_steps
```

The interval which contains the maximum number of steps is `r max_steps[1]`

## Imputing missing values

Calculate the number of missing values
```{r}
total_na <- sum(is.na(activity_data$steps))
total_na
```

There are `r total_na` missing values in the dataset.

### Replace the NA's with the average number of steps for each time interval

Add a column with the average number of steps for each interval
```{r}
add_avg_steps <- activity_data %>% left_join(intervals, by = "interval")
add_avg_steps
```

Create a column `new_steps` that replaces missing values with the average number of steps for that interval
```{r}
replace_nas <- add_avg_steps %>%
  mutate(new_steps = if_else(is.na(steps), avg_steps, steps))
```

Select only columns that are needed
```{r}
activity_data_upd <- replace_nas %>%
  select(new_steps, date, interval) %>%
  rename(steps = new_steps)
activity_data_upd
```

Group the data by day. Find the total steps taken on each day
```{r}
steps_per_day_upd <- activity_data_upd %>%
  group_by(date) %>%
  summarise(total_steps = sum(steps))
steps_per_day_upd
```

Construct a histogram with the updated data
```{r histogram_upd}
hist(steps_per_day_upd$total_steps,
     col = "green",
     main = "Total Number of Steps Taken per Day",
     xlab = "Number of steps",
     ylab = "Frequency")
```

Calculate the mean and median number of steps taken in a day
```{r}
mean_steps_upd <- mean(steps_per_day_upd$total_steps)
median_steps_upd <- median(steps_per_day_upd$total_steps)
```

The mean number of steps taken per day is `r mean_steps_upd` steps.  
The median number of steps taken per day is `r median_steps_upd` steps.  

The mean and the median are now equal.  
The mean has increased slightly, but the median is unchanged.

## Are there differences in activity patterns between weekdays and weekends?

Add a column that indicates whether a day is a weekday or a weekend
```{r}
activity_weekday <- activity_data_upd %>%
  mutate(Day = weekdays(as.Date(date))) %>%
  mutate(Day_type = if_else(Day %in% c("Saturday", "Sunday"), "weekend", "weekday")) %>%
  select(-Day)
activity_weekday
```

Convert `Day_type` into a factor
```{r}
activity_weekday$Day_type <- as.factor(activity_weekday$Day_type)
activity_weekday
```

Find the average number of steps taken at each interval
```{r}
intervals_weekday <- activity_weekday %>%
  group_by(interval, Day_type) %>%
  summarise(avg_steps = mean(steps))
intervals_weekday
```

Construct a panel plot containing a time series plot of the 5-minute interval and the average number of steps taken, averaged across all weekday days or weekend days
```{r panelplot}
xyplot(avg_steps ~ interval | Day_type, data = intervals_weekday, layout = c(1,2),
     type = "l",
     col = "blue",
     main = "Average Number of Steps Taken during each Interval",
     xlab = "Interval",
     ylab = "Number of steps")
```

